Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance in zooplankton

### 2023-05-11 Manuscript/Manuscript.tex
Update Key resource table

The ability to perceive changes in the environment and adapt behaviour
accordingly is essential for the survival of all living organisms.
Avoidance from UV radiation, which damages DNA essential for survival,
is proposed to be one of the behaviours that existed before the abundant
development of Earth's oxygen to block UV radiation in the atmosphere
(Pittendrigh). In the ocean, almost all zooplankton have a pronounced
rhythmic behaviour called Diel vertical migration (DVM), which consists
of upwards towards the surface at dusk and descend back to deeper water
before dawn. Avoidance behaviour to escape harmful UV radiation during
the day is closely linked to the control of vertical migration in the
ocean. However, little is known about the molecular neural circuit
mechanisms in UV avoidance behaviour in many animals, including
zooplankton. In neural circuits in general light response, the
activation of various receptors by excitatory and inhibitory
neurotransmitters such as glutamate, GABA and dopamine is known in
various species. In addition to this, it has long been known that in
nitric oxide (NO), as a neuromodulatory molecule, it is also important
as a modulator of visual activity. NO is a small free-radical molecule
with diffusible and changeable characteristics, and thus works as one of
the neural signalling molecules. The function of NO in the neural
circuitry of vision has been well studied in vertebrates. In the
mammalian retina, NO is an essential signalling molecule in all retinal
cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one
of the retinal cells, activates \emph{soluble guanylate cyclase (sGC)}
expressed in cone bipolar cells, which produces cGMP and regulates
downstream signalling. We therefore hypothesise that NO-mediated neural
signalling may be involved in the photoreceptor circuits that control UV
avoidance behaviour. To test this hypothesis we used the annelid
\emph{Platynereis dumerilii} \textbf{(Figure 1A)}, which conserved
vertebrate-type ciliary photoreceptors (cPRCs) (Arendt et al., 2004).
The cPRC expresses \emph{c-opsin1}, which absorbs UV radiation, is
involved in melatonin production and regulates circadian swimming
behaviour (Tosches et al., 2014; Tsukamoto et al., 2017). Furthermore,
we previously reported abnormal UV avoidance behaviour in
\emph{c-opsin1} knockout larvae (Verasztó et al., 2018). The cPRCs
express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae
(Verasztó et al., 2018). In adult worms, UV light and c-opsin1 modulate
locomotion and maturation (Veedin Rajan et al., 2021). These suggest
that Platynereis larvae are a highly suitable model for elucidating the
neural circuit mechanisms of NO-induced UV avoidance behaviour, and the
accessibility of genome editing, behavioural assays, calcium imaging and
whole-body connectomics could be very useful for study at the molecular
level(Ozpolat et al., 2021).

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of metazoan NOS sequences
\textbf{(Figure 1---figure supplement 1)}. To characterise the
expression pattern of \emph{NOS} we used in situ hybridization chain
reaction (HCR) and transient transgenesis. In two- and three-day-old
larvae, we detected \emph{NOS} expression in four cells (two of them
weakly expressing) in the apical organ region \textbf{(Figure 1D and
Figure 1---figure supplement 2)}. The same four cells were labelled with
a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream
genomic region of the \emph{NOS} gene \textbf{(Figure 1E)}. This
reporter also revealed the axonal projections of the NOS-expressing
neurons. The position and morphology of the four \emph{NOS}-expressing
cells allowed us to identify the same cells in the three-day-old
whole-body volume EM data (Verasztó et al., 2020; Williams et al., 2017)
as four interneurons (INNOS) \textbf{(Figure 1B,C)}. The INNOS cells are
postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW
interneurons, which are also cPRC targets \textbf{(Figure 1C,F)}. INRGW
neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1),
which synapse on the prototroch ciliary band and the cholinergic MC
ciliomotor neurons \textbf{(Figure 1C,F)}. We also detected \emph{NOS}
expression by HCR in the region of the visual eyes (adult eyes) and the
pigmented eyespots \textbf{(Figure 1---figure supplement 1)}.

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** SEM image of a three-day-old Platynereis larva. **(B, C)** Five types of neurons (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a three-day-old larva. Ciliated cells labelled with (B) are shown in grey. The INNOS on the ventral side labelled with (C) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. **(D)** Expression analysis of the NOS gene (magenta) using in situ HCR. Anterior view of the larva at two-day-old. Co-stained image using acetylated α-tubulin antibody (acTub: green). **(E)** Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (NOSp::palmi-3xHA: magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia of cPRC and axonal scaffold (acTub: green). Views of larvae at three-day-old from the anterior side. Insert: One of the reconstructed INNOS is shown from the anterior side. **(F)** Wiring diagram of the cPRC circuit. Hexagons and arrows represent cell groups and their synaptic connections, respectively. Numbers inside hexagons indicate the number of cells grouped in each node. Arrow line thickness (synapse weight) is equal to the number of synapses.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

The specific expression of \emph{NOS} in interneurons directly
postsynaptic to the cPRCs suggests that NO signalling is involved in the
UV response. To test if NO is produced during UV/violet responses, we
injected zygotes with the fluorescent NO-reporter DAF-FM and imaged
two-day-old larvae during the local stimulation of the cPRC cilia by 405
nm light. Following light stimulation, we detected an increase in DAF-FM
fluorescence in the region of the INNOS cells \textbf{(Figure 2)}.

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{../figures/Fig2.png}
\caption{\textbf{Figure 2. NO produced by UV/violet stimulation to
cPRCs.} \textbf{(A)} A NO-positive cell visualised by DAF-FM (green) is
shown (dashline). The white line indicates the outer frame of the larva.
All subsequent images are of two-day-old larvae. \textbf{(B)} Intensity
changes before (upper) and during 405 nm light stimulation (lower) in a
cell visualised by DAF-FM. \textbf{(C)} The intensity of DAF-FM changes
over time when cPRC (green) and other cells (ctr: gray) are stimulated
with UV/violet (405 nm) light. Purple squares indicate the timing of 405
nm light stimulation, the transparent lines show the optical responses
to individual intensities, each intensities are normalized (ΔF/F0) and
the dark trace lines and the shaded bands show the mean value and the
standard error range of the mean.}
\end{figure}

@@

To test this, we generated two \emph{Platynereis} \emph{NOS} knockout
lines with the CRISPR/Cas9 method. We recovered two deletions
(NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus
likely representing null alleles \textbf{(Figure 3A)}. We could
establish a homozygous line for both mutations indicating that NOS is
not an essential gene. To quantify UV avoidance, we recorded the
trajectories of freely swimming larvae in vertical columns, illuminated
laterally from two opposite sides with 395 nm UV/violet light. To
analyse the behaviour of larvae before and after UV stimulation, the
temporal changes in some parameters were compared using tracking of
individual larvae of the WT and mutant \textbf{(Figure 3B,C and Figure
3---figure supplement 1A-D)}. The results showed that in the WT there
was a rapid and strong increase in downward swimming approximately 15 s
after UV stimulation, whereas this did not occur in the mutant
\textbf{(Figure 3C and Figure 3---figure supplement 1B)}. Interestingly,
there was already a difference in swimming speed between the WT and the
mutant before UV stimulation (in the dark) \textbf{(Figure 3---figure
supplement 1C,D)}. After UV stimulation, the swimming speed of the WT
increased and reached a maximum after about 20 s. In contrast, the
mutants showed a slight increase in speed after UV stimulation, but
swimming speed was not maintained and returned to the speed before UV
stimulation. We also tested phototaxis, by exposing larvae to 480 nm
directional and collimated light from the top of the column. As
previously shown, wild-type larvae swim downward within 30 sec following
non-directional UV/violet light stimulation (Verasztó et al., 2018). In
contrast, homozygous \emph{NOS}-mutant larvae showed a strongly
diminished UV-avoidance response \textbf{(Figure 3D and Figure
3---figure supplement 1E)}. This phenotype is similar to the defective
UV-avoidance of \emph{c-opsin1} mutant larvae and reveals a requirement
for NOS in UV-avoidance behaviour (Verasztó et al., 2018).
\emph{NOS}-mutant larvae also showed reduced phototactic behaviour,
suggesting a function in the visual eyes mediating phototaxis (Randel et
al., 2014), where \emph{NOS} is also expressed \textbf{(Figure
1---figure supplement 1A,B)}. To test if the observed defects are due to
a role of NOS in neuronal signalling or circuit developmental, we also
tested larvae exposed to a NOS inhibitor, L-NAME. Larvae incubated for 5
min in 0.1 mM or 1 mM L-NAME showed defective UV avoidance, but normal
phototaxis \textbf{(Figure 3E)}. Overall, our results indicate an acute
requirement for NOS signalling in UV-avoidance and a possible indirect,
developmental role in the visual system. To investigate the role of NO
on UV avoidance behaviour, receptors for NO were explored.

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{../figures/Fig3.png}
\caption{\textbf{Figure 3. Larvae with inhibited NOS have a delayed
response to UV-avoidance.} \textbf{(A)} Top: The domain and Exon/Intron
structure of NOS. Bottom: Close-up region showing the genomic locus of
NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The
generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target
sites. Gray shows PAM sequences, red shows stop codons. \textbf{(B)}
Overlaid trajectories for WT (n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and
NOSΔ23/Δ23, n=47) at three-day-old larvae. 0 sec as the starting point.
After 10 sec, UV (395 nm) stimulation from the side. \textbf{(C)} The
temporal changes in the vertical position of the WT and mutant larvae
before (dark) and after UV stimulation (side 395 nm) are shown. The
starting points of each larval trajectory are set to 0. \textbf{(D)}
Vertical swimming in wild-type (WT) and mutant (NOSΔ11 and NOSΔ23)
larvae at three-day-old stimulated with UV (395 nm) light from side,
blue (488 nm) light from top and UV (395 nm) light from top. The data
are shown in 30 s bins. \textbf{(E)} Vertical swimming in larvae treated
with NOS inhibitors, L-NAME at three-day-old stimulated with UV (395 nm)
light from side, blue (488 nm) light from top and UV (395 nm) light from
top. The data are shown in 30 s bins.}
\end{figure}

@@

NO generally acts via soluble guanylate cyclases (sGC), belonging to the
guanylate cyclase family with a CYC domain (PF00211). NO binding to the
heme group of sGC leads to increased cyclic guanosine monophosphate
(cGMP) production. Analysis of sGCs in \emph{Platynereis} indicated that
these genes are not expressed in any of the cells of the cPRC circuit
(Verasztó et al., 2017). Recently, Moroz and coworkers reported an
atypical but widely conserved family of guanylyl cyclases with a NIT
(nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential
mediators of NO signalling (Moroz et al., 2020). To search for NIT-GCs
in \emph{Platynereis}, we searched transcriptome resources and retrieved
15 potential NIT-GC homologs \textbf{(Figure 4---figure supplement 1)}.
To analyse the relationship of these sequences to metazoan NIT-GCs, we
retrieved protein sequences with a CYC domain from the transcriptome and
genome databases of 45 metazoan species and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a
group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In
agreement with Moroz et al. (Moroz et al., 2020),we found a group of GCs
with NIT domains with representatives in placozoans, cnidarians, some
ecdysozoans, echinoderms, and lophotrochozoans. The 15
\emph{Platynereis} sequences were parts of several major and ancient
clades of the tree \textbf{(Figure 4---figure supplement 1)}. To
characterise the expression of NIT-GCs, we using previously published
spatially mapped single-cell transcriptome data (Achim et al., 2015;
Williams et al., 2017). Among the 15 NIT-GCs, two showed high and
specific expression in the cPRCs and one was expressed in the INNOS
cells \textbf{(Figure 4---figure supplement 1)}. In the single-cell
data, we could identify the cPRCs by the specific expression of c-opsin1
and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide)
(Arendt et al., 2004; Williams et al., 2017), which have previously been
described as cPRC markers \textbf{(Figure 4---figure supplement 2A)}.
The INNOS cells were identified by NOS expression and their spatial
mapping. We decided to focus on two NIT-GCs expressed in the cPRCs and
with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the
single-cell data, we first carried out in situ HCR with probes for
\emph{NIT-GC1} and \emph{NIT-GC2} mRNA. Both genes showed very clear and
specific expression in the four cPRCs, as confirmed by co-labeling with
an acetylated α-tubulin antibody and with an HCR probe against \emph{MLD
proneuropeptide} \textbf{(Figure 4A,B and Figure 4---figure supplement
2A-C)}. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2
at the protein level, we raised and affinity-purified polyclonal
antibodies against a specific peptide sequence from both proteins. In
immunostainings, we found that NIT-GC1 was localise to the region
corresponding to the axonal projections of the cPRCs in the anterior
nerve plexus \textbf{(Figure 4C)}. Co-immunostaining with the rabbit
NIT-GC1 and the rat NOS antibodies revealed that both proteins were
localised in close proximity in the neurosecretory plexus
\textbf{(Figure 4---figure supplement 2D)}. In contrast, NIT-GC2
strongly and specifically labelled the large sensory ciliary region of
the cPRCs \textbf{(Figure 4D)}. These different subcellular
localisations suggest that the two NIT-GCs are involved in different
intracellular signalling processes in the ciliary and axonal regions of
the cPRCs.

\textbf{NIT-GC1 is required for the depolarising response of cPRC to UV
stimulation}

To investigate how NIT-GCs-mediated NO signalling is involved in cPRC
activation, the calcium sensor GCaMP6s was ubiquitously expressed in
larvae and calcium imaging in the cPRC during UV/violet stimulation. As
we have shown previously, a 20-sec local stimulation of the ciliated
part of the cPRC with 405 nm light lead to a transient decrease in cPRC
calcium levels, likely due to the hyperpolarisation of the cPRCs
(Verasztó et al., 2018). After 20-sec, calcium levels in cPRCs were
raising again, reaching higher levels than at the start of the stimulus
- a response that may involve depolarisation. This activation phase
occurs after the end of the stimulation and may be due to a slower
neuroendocrine feedback (Verasztó et al., 2018). First, to determine
whether NO mediates such a feedback, we measured calcium levels in
\emph{NOS}-mutant larvae. While we detected the inhibition phase in cPRC
in \emph{NOS}-mutants, this was not followed by activation and calcium
levels remained low \textbf{(Figure 4E,F)}. These results revealed an
essential requirement for NO signalling in the second phase of the cPRC
response. Next, to characterise the role of NIT-GCs in changes of
calcium levels in cPRC, calcium imaging was performed when each
\emph{NIT-GC} gene was knocked down using morpholino \textbf{(Figure
4---figure supplement 2E,F)}. In NIT-GC1 morphant larvae, 405 nm light
stimulation caused a decrease in calcium levels in the cPRC, but the
increase in calcium levels that occurred after the light stimulation was
stopped was strongly suppressed \textbf{(Figure 4G)}. This result is
similar to the response of NOS mutants in cPRC, suggesting that NIT-GC1
mediates depolarising responses after UV stimulation of cPRC by NO and
may be involved in extracellular signalling. Interestingly, in NIT-GC2
morphant larvae, an increase in calcium levels occurred immediately
after light stimulation \textbf{(Figure 4H)}. The different responses of
cPRC upon knockdown of NIT-GC1 and NIT-GC2 suggest that there may be
different intracellular signalling mechanisms associated with their
respective localisations. We focused our analysis on NIT-GC1 because it
was localised to cPRC neuroaxons within the neuroplexas region and is
involved in extracellular signalling after UV stimulation. It is
generally known that depolarisation in the PRC involves elevated cGMPs
in the photoreceptors. We therefore used an in vitro system in which
cultured cells expressed NIT-GC1 to assess the cGMP activity of NIT-GC1.

>>>>

Neuronally produced nitric oxide (NO) can act as a diffusible
neuromodulator. This small free-radical molecule is synthesized by
nitric oxide synthase (NOS)\ldots{}

In the mammalian retina, NO is produced by amacrine cells and activates
a \emph{soluble guanylate cyclase (sGC)} in cone bipolar cells. The
NO-dependent increased in cGMP tunes bipolar cell activity \ldots.
{[}refs{]} {[}other examples{]}

To investigate NO signalling in the modulation of neural circuit
dynamics, we studied the larval stages of the marine annelid model
\emph{Platynereis dumerilii} (Ozpolat et al., 2021). We uncovered an
essential function for NO in these larvae in UV/violet-light avoidance.
In \emph{Platynereis} larvae, UV/violet avoidance is characterised by
downward swimming and is mediated by the UV/violet-responding brain
ciliary photoreceptor cells (cPRCs).

The cPRCs express a vertebrate-type ciliary opsin, c-opsin1 (Arendt et
al., 2004). This opsin forms a UV-absorbing bistable photopigment with
an absorption maximum around 384 nm Verasztó et al. (2018).
\emph{c-opsin1} knockout larvae show defective UV/violet avoidance
(Verasztó et al., 2018). Here we show that \emph{Platynereis} \emph{NOS}
is expressed in interneurons of the cPRC circuitry and is required for
UV/violet-avoidance behaviour. By combining calcium imaging across the
fully-mapped cPRC circuit, genetic perturbations and mathematical
modelling, we describe how NO tunes circuit dynamics through
non-synaptic reverse signalling to cPRCs. This delayed neuroendocrine
feedback is essential for circuit output, and integrates UV/violet input
and also induces a short-term memory manifest in altered circuit
activity.

\#In adult worms, UV light and c-opsin1 modulate locomotion and
maturation (Veedin Rajan et al., 2021).

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of metazoan NOS sequences
\textbf{(Figure 1---figure supplement 1)}. To characterise the
expression pattern of \emph{NOS} we used in situ hybridization chain
reaction (HCR) and transient transgenesis. In two- and three-day-old
larvae, we detected \emph{NOS} expression in four cells (two of them
weakly expressing) in the apical organ region \textbf{(Figure 1D and
Figure 1---figure supplement 2)}. \emph{NOS} was also expressed in the
region of the visual eyes (adult eyes) and the pigmented eyespots
\textbf{(Figure 1---figure supplement 1)}. The four apical organ cells,
but not the eyes, were also labelled with a \emph{NOS} reporter
construct driving palmitoylated-tdTomato \textbf{(Figure 1E)}. This
reporter also revealed the axonal projections of these central
NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017)
\textbf{(Figure 1B,C)}. In the synaptic connectome, the INNOS cells are
postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW
interneurons, which are also cPRC targets \textbf{(Figure 1C,F)}. INRGW
neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1),
which synapse on the prototroch ciliary band and the cholinergic MC
ciliomotor neurons \textbf{(Figure 1C,F)}.

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** SEM image of a three-day-old Platynereis larva. **(B, C)** Five types of neurons (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a three-day-old larva. Ciliated cells labelled with (B) are shown in grey. The INNOS on the ventral side labelled with (C) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. **(D)** Expression analysis of the NOS gene (magenta) using in situ HCR. Anterior view of the larva at two-day-old. Co-stained image using acetylated α-tubulin antibody (acTub: green). **(E)** Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (NOSp::palmi-3xHA: magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia of cPRC and axonal scaffold (acTub: green). Views of larvae at three-day-old from the anterior side. Insert: One of the reconstructed INNOS is shown from the anterior side. **(F)** Wiring diagram of the cPRC circuit. Hexagons and arrows represent cell groups and their synaptic connections, respectively. Numbers inside hexagons indicate the number of cells grouped in each node. Arrow line thickness (synapse weight) is equal to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the ramified sensory cilia of the cPRCs to 405 nm violet light.
Following light stimulation of the cPRCs but not a control area, we
detected an increase in DAF-FM fluorescence in the anterior
neurosecretory neuropil, the region of INNOS projections \textbf{(Figure
2)}.

\begin{figure}
\includegraphics[width=29.17in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** A NO-positive cell visualised by DAF-FM (green) is shown (dashline). The white line indicates the outer frame of the larva. All subsequent images are of two-day-old larvae. **(B)** Intensity changes before (upper) and during 405 nm light stimulation (lower) in a cell visualised by DAF-FM. **(C)** The intensity of DAF-FM changes over time when cPRC (green) and other cells (ctr: gray) are stimulated with UV/violet (405 nm) light. Purple squares indicate the timing of 405 nm light stimulation, the transparent lines show the optical responses to individual intensities, each intensities are normalized (ΔF/F0) and the dark trace lines and the shaded bands show the mean value and the standard error range of the mean.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

Next, we wanted to directly test whether NO signalling is involved in
UV/violet avoidance. To achieve this, we generated two
\emph{Platynereis} \emph{NOS} knockout lines with the CRISPR/Cas9
method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both
leading to frame-shift and an early stop codon and thus likely
representing null alleles \textbf{(Figure 3A)}. We could establish a
homozygous line for both mutations indicating that NOS is not an
essential gene. To quantify UV avoidance, we recorded the trajectories
of freely swimming wild type and mutant larvae in vertical columns,
illuminated laterally from two opposite sides with 395 nm UV light
\textbf{(Figure 3B, C and Figure 3---figure supplement 1A-D)}. As
previously shown, wild-type larvae swim downward following
non-directional UV/violet light stimulation (Verasztó et al., 2018). In
contrast, both two- and three-day-old homozygous \emph{NOS}-mutant
larvae showed a strongly diminished UV-avoidance response
\textbf{(Figure 3B-D and Figure 3---figure supplement 1)}. This
phenotype is similar to the defective UV-avoidance of \emph{c-opsin1}
mutant larvae and reveals a requirement for NOS in UV-avoidance
behaviour (Verasztó et al., 2018). Wild type but not mutant larvae also
showed an increase in swimming speed under UV light that may be due to
downward swimming trajectories (swimming with gravity) \textbf{(Figure
3---figure supplement 1C,D)}. We also tested directional phototaxis, by
exposing larvae to 480 nm directional and collimated light from the top
of the column. Three-day-old but not two-day-old \emph{NOS}-mutant
larvae also showed reduced phototactic behaviour, suggesting a function
for \emph{NOS} in the visual eyes that mediate three-day-old phototaxis
(Randel et al., 2014) \textbf{(Figure 1---figure supplement 1A,B)}. To
distinguish between an acute or developmental function of NOS in light
responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance. In contrast, phototaxis was
not effected \textbf{(Figure 3E)}. Overall, our results indicate an
acute requirement for NOS signalling in UV-avoidance and a possible
indirect, developmental role in visual phototaxis.

\begin{figure}
\includegraphics[width=33.33in]{figures/Fig3} \caption{**Figure 3. Larvae with inhibited NOS have a delayed response to UV-avoidance.** **(A)** Top: The domain and Exon/Intron structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Gray shows PAM sequences, red shows stop codons. **(B)** Overlaid trajectories for WT (n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) at three-day-old larvae. 0 sec as the starting point. After 10 sec, UV (395 nm) stimulation from the side. **(C)** The temporal changes in the vertical position of the WT and mutant larvae before (dark) and after UV stimulation (side 395 nm) are shown. The starting points of each larval trajectory are set to 0. **(D)** Vertical swimming in wild-type (WT) and mutant (NOSΔ11 and NOSΔ23) larvae at three-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins. **(E)** Vertical swimming in larvae treated with NOS inhibitors, L-NAME at three-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins.}\label{fig:unnamed-chunk-3}
\end{figure}

\textbf{NO feedback tunes cPRC responses to UV/violet stimulation}

To investigate how NO signalling alters the dynamics of the cPRC
circuit, we carried out calcium imaging experiments. We ubiquitously
expressed the calcium sensor GCaMP6s in larvae and imaged calcium
signals during 405 nm stimulation of the cPRCs. As we have shown
previously, a 20-sec local stimulation of cPRC cilia lead to a transient
increase in cPRC calcium levels, followed by a transient decrease
(Verasztó et al., 2018). After \textasciitilde20-sec, calcium levels in
cPRCs were raising again, reaching higher levels than at the start of
the stimulus -- a response that may involve depolarisation. This
activation phase occurs after the 20 sec stimulation period and is
likely due to a delayed neuroendocrine feedback (Verasztó et al., 2018).
To determine whether NO mediates such a feedback, we repeated the
experiment in \emph{NOS}-mutant larvae. While we detected the initial
activation phase followed by inhibition, in \emph{NOS}-mutants this was
not followed by delayed activation and calcium levels dropped to a low
steady-state level \textbf{(Figure 4E,F)}. We thus identified a
requirement for NO signalling in the late-phase activation of cPRCs.

@@

Next, we aimed to identify the NO-receptor in the cPRCs. NO generally
acts via soluble guanylate cyclases (sGC), belonging to the guanylate
cyclase family with a CYC domain (PFAM domain: PF00211). NO binding to
the heme group of sGC leads to increased cyclic guanosine monophosphate
(cGMP) production. Analysis of sGCs in \emph{Platynereis} indicated that
these genes are not expressed in any of the cells of the cPRC circuit
(Verasztó et al., 2017). Recently, Moroz and coworkers reported an
atypical but widely conserved family of guanylyl cyclases with a NIT
(nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential
mediators of NO signalling (Moroz et al., 2020). To search for NIT-GCs
in \emph{Platynereis}, we searched transcriptome resources and retrieved
15 potential NIT-GC homologs \textbf{(Figure 4---figure supplement 1)}.
To analyse the relationship of these sequences to metazoan NIT-GCs, we
retrieved protein sequences with a CYC domain from the transcriptome and
genome databases of 45 metazoan species and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a
group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In
agreement with Moroz et al. (Moroz et al., 2020), we found a group of
GCs with NIT domains with representatives in placozoans, cnidarians,
some ecdysozoans, echinoderms, and lophotrochozoans. The 15
\emph{Platynereis} sequences were parts of several major and ancient
clades of the tree \textbf{(Figure 4---figure supplement 1)}. To
characterise the expression of NIT-GCs, we used previously published
spatially mapped single-cell transcriptome data (Achim et al., 2015;
Williams et al., 2017). Among the 15 NIT-GCs, two showed high and
specific expression in the cPRCs and one was expressed in the INNOS
cells \textbf{(Figure 4---figure supplement 1)}. In the single-cell
data, we could identify the cPRCs by the specific expression of c-opsin1
and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide)
(Arendt et al., 2004; Williams et al., 2017), which have previously been
described as cPRC markers \textbf{(Figure 4---figure supplement 2A)}.
The INNOS cells were identified by NOS expression and spatial mapping in
the brain (Achim et al., 2015). We decided to focus on two NIT-GCs
expressed in the cPRCs and with a full-length sequence, NIT-GC1 and
NIT-GC2. To confirm the single-cell data, we first carried out in situ
HCR with probes for \emph{NIT-GC1} and \emph{NIT-GC2} mRNA. Both genes
were specifically expression in the four cPRCs, as confirmed by
co-labeling with an acetylated α-tubulin antibody and with an HCR probe
against \emph{pedal peptide 2/MLD proneuropeptide} \textbf{(Figure 4A,B
and Figure 4---figure supplement 2A-C)}. To analyse the subcellular
localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and
affinity-purified polyclonal antibodies against a specific peptide
sequence from both proteins. In immunostainings, we found that NIT-GC1
was localise to the region corresponding to the axonal projections of
the cPRCs in the anterior nerve plexus \textbf{(Figure 4C)}.
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat
antibody raised against a fragment of \emph{Platynereis} NOS, revealed
that both proteins were localised in close proximity in the
neurosecretory plexus \textbf{(Figure 4---figure supplement 2D)}. In
contrast, NIT-GC2 specifically labelled the large sensory ciliary region
of the cPRCs \textbf{(Figure 4D)}. These different subcellular
localisations suggest that the two NIT-GCs are involved in different
intracellular signalling processes in the ciliary and axonal regions of
the cPRCs.

\textbf{NIT-GC1 produces cGMP in an NO-dependent manner}

To further characterise these atypical guanylyl cyclases, we focused on
NIT-GC1 and carried out in vitro experiments. In bacteria, NIT domains
are thought to regulate cellular functions in response to intra- or
extracellular nitrate and nitrite. NIT-GC1 has a NIT domain and a highly
conserved cyclase domain that is expected to catalyse cGMP synthesis
\textbf{(Figure 4I)}. The NIT domain may render NIT-GC1 dependent on NO
signals. To test this, we co-expressed the synthetic cGMP indicator
Green cGull (Matsuda et al., 2016) and NIT-GC1 in cultured cells COS-7
(monkey kidney). For balanced expression, we used a single plasmid with
the two open-reading frames separated by the 2A self-cleaving peptide
\textbf{(Figure 4J)}. Application of the NO donor SNAP lead to increased
cGMP levels, an effect we did not observe when cells were exposed to
DMSO or when Green cGull was expressed alone \textbf{(Figure 4K-M)}. To
test whether cGMP production is dependent on the NIT domain, we also
tested a deletion construct of NIT-GC1 lacking the NIT domain
\textbf{(Figure 4I)}. Cells expressing this construct and Green cGull
did not show increased cGMP levels when exposed to SNAP \textbf{(Figure
4N)}. These results indicate that NIT-GC1 is able to catalyse cGMP
production in an NO-dependent manner and this function requires the NIT
domain. These results establish NIT-GC1 as a biochemical NO sensor.

\textbf{NIT-GC1 is required for NO-mediated feedback to cPRCs during the
UV response}

To test the in vivo function of NIT-GC1 and NIT-GC2 in cPRC responses,
we combined calcium imaging with morpholino-mediated knockdowns. We used
two translation-blocking morpholinos for each \emph{NIT-GC} gene and
tested knockdown efficiency by immunostaining injected animals with the
NIT-GC1 and NIT-GC2 antibodies \textbf{(Figure 4---figure supplement
2E,F)}. For both genes, the morpholinos led to a strong reduction in the
respective antibody signal, confirming efficient knockdown and antibody
specificity.

In NIT-GC1 morphant larvae, the delayed activation of cPRCs following
405 nm stimulation did not occur \textbf{(Figure 4G)}. This phenotype is
similar to the phenotype of \emph{NOS} mutants suggesting that NIT-GC1
acts as the NO sensor in cPRCs to drive their delayed activation. This
could occur via increased cGMP production and the opening of a
cyclic-nucleotide-gated (CNG) channel specific to cPRCs. NIT-GC2
morphant larvae, in contrast, showed a step-up increase in calcium
following light stimulation \textbf{(Figure 4H)}. The calcium signal
decayed during stimulation and was off after light off. These data
support an essential role for ciliary-localised NIT-GC2 in suppressing
cPRC calcium following its transient rise at stimulus onset. Overall,
these knockdown experiments revealed different signalling mechanisms for
the two NIT-GCs that may be due to their different subcellular
localisations.
